Rationality, Evolution \& the Environment

### 2015-09-15 paper/paper.tex
rewrite section 2

\usepackage[]{helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
\bibliography{helpers/MyRefLocal}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

\title{Rationality, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke}
\date{current version: August 21st 2015}

@@

\section{Rationality, subjectivity  \& subjective regret}
\label{sec:rati--subj}

\todo[inline]{would it make sense to tighten and restructure this section a little bit? what
  about this: rationality says \emph{if} utils, beliefs and rational, then choose this; utils
  and beliefs are unobservable; in some case utils and preferences \emph{can} be tight to the
  environment (probability of a fair coin landing heads), in others no (probability of life on
  Mars); if tight to environment its frequencies and fitness that would pin down a norm for
  rational behavior, otherwise there's the danger of redescriptionism (just find some belief
  and preference that rationalizes observed behavior); \dots }

The standard definition of \textit{rationality} in economics and decision theory says that
choice is rational only if it maximizes (subjective) expected utility. Since the work by vNM
[cite], this view has been justified from the ``intuitive rationality'' of the axioms on the
subjective preference relation leading to the representability of the decision maker as an
expected utility maximizer. \todo{include axioms, or at least those that are mentioned as
  relevant later} While in the axiomatization by [vNM] the probability of the outcomes is
objectively given a priori, Savage54 provides a larger set of axioms that represent the
decision maker (DM) with both a subjective utility over outcomes \textit{and} a
\textit{subjective} probability over states of the world. Savage's axioms constitute the
foundation of subjective expected utility (SEU) theory, that is normally assumed to define the
rational behavior of economic agents.

>>>>

\usepackage[]{helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
% \bibliography{helpers/MyRefLocal}
\bibliography{MyRefGlobal}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

\title{Rationality, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke}
\date{current version: September 15th 2015}

@@

\section{Rationality, subjectivity, beliefs \& preferences}
\label{sec:rati--subj}

The standard textbook definition of \textit{rationality} in economics and decision theory
traces back to the seminal work of \citet{Neumannvon-NeumannMorgenstern1944:Theory-of-Games}
and \citet{Savage1954:The-Foundations}. It says that choice is rational only if it maximizes
(subjective) expected utility.

\begin{definition}[Rational choice from maximization of expected utility] Let $S$ be a (finite)
  set of states of the world and $A$ a (finite) set of actions. Given a probability
  distribution $P$ over $S$ and a utility function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is a rational choice
  only if it maximizes the \emph{(subjective) expected utility}
  $\EU(a^*) = \sum_{s \in S} P(s) \ u(s, a^*)$: $a^* \in \text{argmax}_{a_i \in A} \EU(a^*)
  $. \todo{Give finite formulation? Everything we do here is finite}
\end{definition}

\noindent Expected utility is subjective in the sense that it is a function of subjective
preferences and subjective beliefs of the decision maker (DM). To wit, a choice can be
rational, i.e., the best choice from the DM's point of view, even if based on peculiar beliefs
and/or aberrant preferences. (Indeed, while
\citeauthor{Neumannvon-NeumannMorgenstern1944:Theory-of-Games} assumed that probabilities were
objectively given, it was \citet{Savage1954:The-Foundations} who famously extended von Neumann
and Morgenstern's axiomatization to include a subjective notion of probabilistic beliefs on top
of the subjective notion of preference that von Neumann and Morgenstern started out with.)

If beliefs and preferences are subjective, there is room for \emph{rationalization} or
\emph{redescriptionism} of observable behavior. Prospect theory is, perhaps, the most prominent
example. \citet{KahnemannTversky1979:Prospect-Theory} looked at experimental data from human
choices in suitably defined tasks and demonstrated that what appears to be violations of
rationality norms can be explained as rational on the assumption that subjects' beliefs and
preferences deviate systematically from the objectively given parameters in the presented
choice task (e.g., by taking into account reference points and diminishing returns for both
gains and losses). Similarly, tinkering with the notion of subjective utility, in particular by
including social preferences such as fairness, reciprocity or similar, allows us to describe as
rational empirically observed behavior in social dilemmas, such as the prisoners' dilemma or
public goods games, that might otherwise appear irrational (e.g. Fehr and Schmidt, Charness and
Rabin, etc.).\todo{@Paolo: please include proper references}

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational everything that is observed, given the
freedom to adjust beliefs and preferences freely. Of course, there are many ways in which
ascription of beliefs and preferences is constrained further by normative considerations of
rationality. \todo{speak about Dutch books here?} To begin with, if the DM's uncertainty is
about objective chance, it can be argued that subjective beliefs should conform (at least
roughly) to objective chance. Similarly, in certain situations it may make sense to postulate
on normative grounds that subjective preferences should be oriented towards tracking objective
utility. For instance, in their classic work \myalert{(Alchian, Friedman)} argue that profit
maximization is a reasonable assumption for characterizing outcomes in a competitive market
because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run.\todo{@Paolo: please add references}

An alternative to descriptive vs.~normative views on rationality is \emph{adaptationism}
\citep[e.g.][]{HagenChater2012:Decision-Making}. \todo{more references} Adaptationism aims to
explain choice behavior by appeal to evolutionary considerations. The key idea is that we
should see decision makers (human and other life forms) as having evolved mechanisms for
choosing that were fitness enhancing (in the relevant domains of choice making, at the relevant
time of evolution). When it comes to preferences, this is clearly related to Alchian and
Friedman's argument that firms \emph{should} optimize that which makes them ``live,'' so to
speak. For biological agents, the argument would be that subjective preferences \emph{should}
track whatever determines objective fitness.

Taking inspiration from recent economic literature on the evolution of preference, we would
like to put this latter idea to the test. We

\bigskip

 Since the work by vNM [cite], this view has been justified from
the ``intuitive rationality'' of the axioms on the subjective preference relation leading to
the representability of the decision maker as an expected utility maximizer. \todo{include
  axioms, or at least those that are mentioned as relevant later} While in the axiomatization
by [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a
larger set of axioms that represent the decision maker (DM) with both a subjective utility over
outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's
axioms constitute the foundation of subjective expected utility (SEU) theory, that is normally
assumed to define the rational behavior of economic agents.
